\documentclass[11pt]{artikel3}

\usepackage{pslatex,hyperref}

\def\n{\bgroup\catcode`\_=12 \catcode`\~=12 \tt \let\next=}

\addtolength\textwidth{1in}
\addtolength\textheight{-3in}
\addtolength\oddsidemargin{-.5in}

\begin{document}
\title{Instructions for the lab exercises}
\author{}\date{}\maketitle

\section*{Session setup}

\begin{itemize}
\item Log in to your account on longhorn.
\item Do \n{module load petsc/3.2-debug}. If you log out, you have to
  repeat this when you log in again.
\end{itemize}

\section*{Download materials}

\begin{itemize}
\item Download the course materials by
  \n{wget http://tinyurl.com/ISTC-petsc-tutorial} You do not have to do
  this in an empty directory.
\item Expand the file you have just downloaded: \n{tar fxz
  petsc_tutorial_files.tgz} This creates a directory
  \n{petsc_tutorial_files} Go there.
\item You will find: 
  \begin{itemize}
  \item four source directories: exercises and solutions for C and
    Fortran,
  \item files with the lecture slides and the instructions for the
    exercises, and this handout.
  \end{itemize}
\item There is also a job file to submit parallel runs. However, we
  will use a different way of running the exercises.
\end{itemize}

\section*{Interactive development}

\begin{itemize}
\item Create a directory \n{idev}, for instance in your home
  directory.
\item Go there: \n{cd ~/idev}
\item Do: \n{tar fxv /tmp/idev.tar}
\item Now you can start the idev environment with \n{~/idev/idev} or
  just \n{idev} after you log out and log in again.
\item This will give you 8 processors (on longhorn); for lower numbers
  do \n{idev 3 8} et cetera.
\item You can now run programs by \n{ibrun ./program}
\item Unfortunately, you can not compile PETSc programs in this
  environment, so keep an extra window open for doing the compilation.
\item If you already had a TACC account (meaning you didn't create one
  just for this course) then you can specify a further option so that
  your running here does not get charged to your account.

  If you signed up through XSEDE, do \n{idev -A TG-TRA120007}

  If you signed up through the TACC website, do \n{idev -A 20120117PETSC}
\end{itemize}

\end{document}
