Here are some guidelines for how to submit assignments and projects.
As a general rule, consider programming as an experimental science,
and your writeup as a report on some tests you have done: explain
the problem you're addressing, your strategy, your results.

\paragraph*{\bf Structure of your writeup}

Most of the projects in this book use a scientific question to allow
you to prove your coding skills.
That does not mean that turning in the code is sufficient, nor code plus sample output.
Turn in a writeup in pdf form that was generated from a text
processing program such 
(preferably) \LaTeX\ (for a tutorial, see~\HPSCref{tut:latex}).

Your writeup should have:
\begin{itemize}
\item Foremost, a short description of the purpose of your project and your results;
\item An explanation of your algorithms or solution strategy;
\item Relevant fragments of your code;
\item A scientific discussion of what you observed,
\item Any code-related observations.
\item If applicable: graphs, both of application quantities and
performance issues.
(For parallel runs possibly TAU plots; see~\ref{tut:tau}).
\end{itemize}

\paragraph*{Observe, measure, hypothesize, deduce}

Your project may be a scientific investigation of some
phenomenon. Formulate hypotheses as to what you expect to observe,
report on your observations, and draw conclusions.

Quite often your program will display unexpected behaviour. It is important to observe
this, and hypothesize what the reason might be for your observed behaviour.

In most applications of computing machinery we care about the efficiency with which
we find the solution. Thus, make sure that you do measurements. In general, make
observations that allow you to judge whether your program behaves the way you
would expect it to.

\paragraph*{Including code}

If you include code samples in your writeup, make sure they look good. For starters,
use a mono-spaced font. In \LaTeX, you can use the \n{verbatim} environment or the 
\n{verbatiminput} command. In that section option the source is included automatically,
rather than cut and pasted. This is to be preferred, since your writeup will
stay current after you edit the source file.

Including whole source files makes for a long and boring writeup. The code samples in this
book were generated as follows. In the source files, the relevant snippet was marked as
\begin{verbatim}
... boring stuff
//snippet samplex
  .. interesting! ..
//snippet end
... more boring stuff
\end{verbatim}
The files were then processed with the following command line (actually, included
in a makefile, which requires doubling the dollar signs):
\begin{verbatim}
for f in *.{c,cxx,h} ; do
  cat $x | awk 'BEGIN {f=0}
                /snippet end/ {f=0}
                f==1 {print $0 > file}
                /snippet/ && !/end/ {f=1; file=$2 }
               '
done
\end{verbatim}
which gives (in this example) a file \n{samplex}. Other solutions are of course possible.

\paragraph*{\bf Code formatting}

Included code snippets should be readable. At a minimum you could 
indent the code correctly in an editor before you include it in
a \indextermtt{verbatim} environment. (Screenshots of your terminal
window are a decidedly suboptimal solution.)
But it's better to use the \indextermtt{listing} package which formats
your code, include syntax coloring. For instance, 
\begin{verbatim}
\lstset{language=C++} % or Fortran or so
\begin{lstlisting}
for (int i=0; i<N; i++)
  s += 1;
\end{lstlisting}
\end{verbatim}
\lstset{language=C++} % or Fortran or so
\begin{lstlisting}
for (int i=0; i<N; i++)
  s += 1;
\end{lstlisting}

\paragraph*{\bf Running your code}

A single run doesn't prove anything. For a good report, you need to
run your code for more than one input dataset (if available) and in
more than one processor configuration. When you choose problem sizes,
be aware that an average processor can do a billion operations per
second: you need to make your problem large enough for the timings to
rise above the level of random variations and startup phenomena.

When you run a code in parallel, beware that on clusters the behaviour
of a parallel code will always be different between one node and
multiple nodes.  On a single node the MPI implementation is likely
optimized to use the shared memory. This means that results obtained
from a single node run will be unrepresentative. In fact, in timing
and scaling tests you will often see a drop in (relative) performance
going from one node to two.  Therefore you need to run your code in a
variety of scenarios, using more than one node.

\paragraph*{Reporting scaling}

If you do a scaling analysis, a graph reporting runtimes should not
have a linear time axis: a logarithmic graph is much easier to
read. A~speedup graph can also be informative.

Some algorithms are mathematically equivalent in their sequential and
parallel versions. Others, such as iterative processes, can take more
operations in parallel than sequentially, for instance because the
number of iterations goes up. In this case, report both the speedup of
a single iteration, and the total improvement of running the full
algorithm in parallel.

\paragraph*{Repository organization}

If you submit your work through a repository, make sure you organize
your submissions in subdirectories, and that you give a clear name to
all files. Object files and binaries should not be in a repository
since they are dependent on hardware and things like compilers.
